---
title: "Simple_Linear_Regression_Walkthrough"
author: "D. Pluta"
date: "2024-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## General Problem Setting

We are given a set of $n$ observations $\{(x_i, y_i)\}_{i = 1}^n$, 
with response $y_i \in \mathbb{R}$ and covariate $x_i$ (the data type 
for the covariate may be numeric or categorical).

Typically, the analysis goal is to determine whether the covariate is significantly 
associated with the reponse as specified by some systematic functional relationship.  It 
is assumed that the observed response values are contaminated by the presence of 
random noise.

## Model Statement

$$\begin{align}
y_i &= \beta_0 + \beta_1 x_i + \varepsilon_i\\
\varepsilon &\sim \mathcal{N}(0, \sigma^2)
\end{align}$$

## Model Assumptions

1. Linear relationship of $x$ with $y$.

2. Independence of error terms.

3. Normally distributed errors.

4. Equal variance of the error distribution over the range of observed covariate values.

## Data Example

```{r}
library(dplyr)
library(ggplot2)
library(readr)
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
dat_raw <- read_csv("C://Users//faculty//Downloads//cancer_reg.csv")
colnames(dat)
```

```{r}
dat %>% select(TARGET_deathRate, medIncome) %>% 
  head()
```


```{r}
fit <- lm(TARGET_deathRate ~ medIncome, data = dat)
summary(fit)
```
```{r}
dat <- dat %>% transmute(TARGET_deathRate = TARGET_deathRate,
                         medIncome10k = medIncome / 10000)
dat$medIncome10k <- dat$medIncome10k - mean(dat$medIncome10k)
```
```{r}
fit <- lm(TARGET_deathRate ~ medIncome10k, data = dat)
summary(fit)
```
```{r}
ggplot(dat) + 
  geom_histogram(aes(x = medIncome10k))
```
```{r}
ggplot(dat) + 
  geom_histogram(aes(x = TARGET_deathRate))
```

```{r}
ggplot(dat) + 
  geom_point(aes(x = medIncome10k, y = TARGET_deathRate))
```

```{r}
ggplot(dat) +
  geom_point(aes(x = medIncome10k, y = TARGET_deathRate), alpha = 0.5, pch = 21, color = "black", fill = "darkgreen") + 
  geom_smooth(aes(x = medIncome10k, y = TARGET_deathRate), method = "lm") + 
  geom_abline(aes(slope = -4.9, intercept = 178.7), color = 'red')
```

```{r}
dat <- dat_raw
```

## Multiple Linear Regression Analysis

Include covariates: gender, race, age
Response: death rate

```{r}
colnames(dat)
dat <- dat_raw %>% select(medIncome, MedianAge, PctWhite, PctBlack, PctAsian, PctOtherRace, MedianAgeMale, MedianAgeFemale, TARGET_deathRate)
```

```{r}
head(dat)
dat$medIncome <- (dat$medIncome  - mean(dat$medIncome)) /10000
```

```{r}
ggplot(dat) + 
  geom_point(aes(x = MedianAge, y = TARGET_deathRate))
```

```{r}
summary(dat$MedianAge)
sum(dat$MedianAge > 100)
```
```{r}
ggplot(dat %>% filter(MedianAge > 100)) + 
  geom_point(aes(x = MedianAge, y = TARGET_deathRate))
```

```{r}
cor(dat$MedianAge[dat$MedianAge > 100], dat$medIncome[dat$MedianAge > 100])
cor(dat$MedianAge, dat$medIncome)
```
```{r}
dat <- dat %>% filter(MedianAge < 100)
```

```{r}
fit <- lm(TARGET_deathRate ~ medIncome + MedianAge, data = dat)
summary(fit)
```
```{r}
par(mfrow = c(1, 1))
plot(fit)
```
```{r}
resids <- fit$residuals
plot(dat$MedianAge, resids)
plot(dat$medIncome, resids)
```

```{r}
dat
X <- as.matrix(dat[, 1:8])
X <- cbind(rep(1, nrow(X)), X)
Y <- dat$TARGET_deathRate
```

```{r}
fit <- lm(TARGET_deathRate ~ ., data = dat)
summary(fit)
```
```{r}
covariance = t(X) %*% X
XTXinv = solve(t(X) %*% X)

beta_hat = XTXinv %*% t(X) %*% Y
```



